### 2009-02-23 htdocs/sections.tex
Correction to surfactant p0
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@471 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  sections.tex
%
%  Content for Ludwig documentation.
%
%  $Id: sections.tex,v 1.8 2009-01-29 12:25:43 kevin Exp $
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  Kevin Stratford (kevin@epcc.ed.ac.uk)
%  (c) 2008 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

A number of different free energies have been used to study
surfactants [cite Yeomans]. Here, the free energy is based on
that used by van der Graff and van der Sman and others [cite].
There are a number of different contributions to the free
energy density:
\begin{equation}
f = f_\phi + f_\psi +f_1 + f_{ex}.
\end{equation}
The first is exactly that used for the binary fluid, i.e.,
\begin{equation}
f_\phi (\phi, \partial_\alpha\phi) =
{\textstyle \frac{1}{2}} A \phi^2 + {\textstyle \frac{1}{4}} B \phi^4
+ {\textstyle \frac{1}{2}} \kappa (\partial_\alpha \phi)^2.
\end{equation}
The surface tension and the interfacial width are determined from the
choice of parameters $A,B,\kappa$ as before.
The second term in the free energu density represents the entropic
energy of mixing of the surfactant with the bulk phases
\begin{equation}
f_\psi (\psi) = kT \left[ \psi \ln\psi + (1 - \psi)\ln(1-\psi) \right]
\end{equation}
where the parameter $kT$ sets the energy scale for the whole system.
The surfactant order parameter, $\psi$, is unity for a saturated
interface and zero for a plain interface.

The surfactant lowers the energy of the interface where it is adsorbed
there, represented by the term
\begin{equation}
F_1(\psi, \partial_\alpha \phi) 
= -{\textstyle \frac{1}{2}} \epsilon \psi (\partial_\alpha \phi)^2
  -{\textstyle \frac{1}{2}} \beta  \psi^2 (\partial_\alpha \phi)^2
\end{equation}
where $\epsilon$ and $\beta$ are parameters. Note that only the
first term appears in \cite{vandergraaf}, while the second term is
based on that appearing in, e.g., \cite{diamant}. If $\beta = 0$,
the surfactant profile at the interface is symmetric (so-called
Langmuir adsoption) whereas the term in $\beta$ induces an
asymmetry (Frumpkin adsorption).
Finally,  $f_{ex}$ is an enthalpic contribution introduced 
to stabilise the interface \cite{theissengompper}
\begin{equation}
f_{ex} (\psi, \partial_\alpha \phi) = {\textstyle\frac{1}{2}} W \psi \phi^2,
\end{equation}
with $W$ a parameter.

@@

The respective chemical potentials are obtained from the functional
derivatives of the free energy $\mu_\phi = \delta F / \delta \phi$ and
$\mu_\psi = \delta F / \delta \psi$. We obtain:
\begin{equation}
\mu_\phi =
A\phi + B\phi^3 -\kappa \partial_\alpha^2 \phi
+ W\psi\phi + \epsilon\psi\partial_\alpha^2 \phi
+ \epsilon \partial_\alpha \phi \partial_\alpha \psi
+ 2\beta \psi\partial_\alpha\phi \partial_\alpha \psi
+ \beta \psi^2 \partial_\alpha^2 \phi,
\end{equation}
and
\begin{equation}
\mu_\psi = kT \left[ \ln\psi - \ln(1-\psi) \right]
+ {\textstyle \frac{1}{2}}W\phi^2
- {\textstyle \frac{1}{2}}\epsilon (\partial_\alpha \phi)^2
- \beta\psi (\partial_\alpha \phi)^2.
\end{equation}
As before, the pressure tensor is $P_{\alpha\beta} = p_0\delta_{\alpha\beta}
+ \tilde{P}_{\alpha\beta}$ with the isotrpic part being determined as (e.g.,
\cite{theissengompper})
\begin{equation}
p_0 = \phi \mu_\phi + \psi \mu_\psi - f(\phi, \psi, \partial_\alpha \phi).
\end{equation}
There is no closed form for the tensor part, which must be choosen to
comply with the constraint that in mechanical equilibrium
\begin{equation}
\partial_\alpha P_{\alpha\beta} = \phi\partial_\alpha\mu_\phi +
\psi\partial_\alpha\mu_\psi.
\end{equation}
In this way we obtain for the isotropic contribution
\begin{eqnarray}
p_0 = {\textstyle \frac{1}{2}}A\phi^2 + {\textstyle \frac{3}{4}}B\phi^4 
- {\textstyle\frac{1}{2}}\kappa (\partial_\alpha\phi)^2 
- \kappa \phi \partial_\alpha^2 \phi
- kT \ln(1-\psi) + W\psi\phi^2
+ \epsilon \phi \partial_\alpha \phi \partial_\alpha \psi
\nonumber\\
+ 2 \beta\phi\psi \partial_\alpha \phi \partial_\alpha\psi
+ \beta \phi \psi^2 \partial_\alpha^2 \phi
- {\textstyle\frac{1}{2}} \beta \psi^2 (\partial_\alpha \phi)^2
\end{eqnarray}
and for the tensor part
\begin{equation}
\tilde{P}_{\alpha\beta} =
(\kappa - \epsilon \psi - \beta\psi^2) 
\partial_\alpha \phi \partial_\beta \phi.
\end{equation}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  sections.tex
%
%  Content for Ludwig documentation.
%
%  $Id: sections.tex,v 1.9 2009-02-23 18:54:09 kevin Exp $
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  Kevin Stratford (kevin@epcc.ed.ac.uk)
%  (c) 2008 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

A number of different free energies have been used to study
surfactants [cite Yeomans]. Here, the free energy is based on
that used by van der Graff and van der Sman and others [cite].
There are a number of different contributions to the free
energy density:
\begin{equation}
f = f_\phi + f_\psi +f_1 + f_{ex}.
\end{equation}
The first is exactly that used for the binary fluid, i.e.,
\begin{equation}
f_\phi (\phi, \partial_\alpha\phi) =
{\textstyle \frac{1}{2}} A \phi^2 + {\textstyle \frac{1}{4}} B \phi^4
+ {\textstyle \frac{1}{2}} \kappa (\partial_\alpha \phi)^2.
\end{equation}
The surface tension and the interfacial width are determined from the
choice of parameters $A,B,\kappa$ as before.
The second term in the free energy density
represents the entropic
energy of mixing of the surfactant with the bulk phases
\begin{equation}
f_\psi (\psi) = kT \left[ \psi \ln\psi + (1 - \psi)\ln(1-\psi) \right]
\end{equation}
where the parameter $kT$ sets the energy scale for the whole system.
The surfactant order parameter, $\psi$, is unity for a saturated
interface and zero for a plain interface.

The surfactant lowers the energy of the interface where it is adsorbed
there, represented by the term
\begin{equation}
f_1(\psi, \partial_\alpha \phi) 
= -{\textstyle \frac{1}{2}} \epsilon \psi (\partial_\alpha \phi)^2
  -{\textstyle \frac{1}{2}} \beta  \psi^2 (\partial_\alpha \phi)^2
\end{equation}
where $\epsilon$ and $\beta$ are parameters. Note that only the
first term appears in \cite{vandergraaf}, while the second term is
based on that appearing in, e.g., \cite{diamant}. If $\beta = 0$,
the surfactant profile at the interface is symmetric (so-called
Langmuir adsoption) whereas the term in $\beta$ induces an
asymmetry (Frumpkin adsorption).
Finally,  $f_{ex}$ is [van der Sman claims]
an enthalpic contribution introduced 
to stabilise the interface \cite{theissengompper}
\begin{equation}
f_{ex} (\psi, \partial_\alpha \phi) = {\textstyle\frac{1}{2}} W \psi \phi^2,
\end{equation}
with $W$ a parameter.

@@

The respective chemical potentials are obtained from the functional
derivatives of the free energy $\mu_\phi = \delta F / \delta \phi$ and
$\mu_\psi = \delta F / \delta \psi$. We obtain:
\begin{equation}
\mu_\phi =
A\phi + B\phi^3 -\kappa \partial_\alpha^2 \phi
+ W\psi\phi + \epsilon\psi\partial_\alpha^2 \phi
+ \epsilon \partial_\alpha \phi \partial_\alpha \psi
+ 2\beta \psi\partial_\alpha\phi \partial_\alpha \psi
+ \beta \psi^2 \partial_\alpha^2 \phi,
\end{equation}
and
\begin{equation}
\mu_\psi = kT \left[ \ln\psi - \ln(1-\psi) \right]
+ {\textstyle \frac{1}{2}}W\phi^2
- {\textstyle \frac{1}{2}}\epsilon (\partial_\alpha \phi)^2
- \beta\psi (\partial_\alpha \phi)^2.
\end{equation}
As before, the pressure tensor is $P_{\alpha\beta} = p_0\delta_{\alpha\beta}
+ \tilde{P}_{\alpha\beta}$ with the isotropic part being determined as (e.g.,
\cite{theissengompper})
\begin{equation}
p_0 = \phi \mu_\phi + \psi \mu_\psi - f(\phi, \psi, \partial_\alpha \phi).
\end{equation}
There is no closed form for the tensor part, which must be chosen to
comply with the constraint that in mechanical equilibrium
\begin{equation}
\partial_\alpha P_{\alpha\beta} = (\phi\partial_\alpha\mu_\phi +
\psi\partial_\alpha\mu_\psi)\delta_{\alpha\beta}.
\end{equation}
In this way we obtain for the isotropic contribution
\begin{eqnarray}
p_0 = {\textstyle \frac{1}{2}}A\phi^2 + {\textstyle \frac{3}{4}}B\phi^4 
- {\textstyle\frac{1}{2}}\kappa (\partial_\alpha\phi)^2 
- \kappa \phi \partial_\alpha^2 \phi
- kT \ln(1-\psi) + W\psi\phi^2
\nonumber\\
+ \epsilon \phi \partial_\alpha \phi \partial_\alpha \psi
+ \epsilon\phi\psi \partial_\alpha^2 \phi
+ 2 \beta\phi\psi \partial_\alpha \phi \partial_\alpha\psi
+ \beta \phi \psi^2 \partial_\alpha^2 \phi
- {\textstyle\frac{1}{2}} \beta \psi^2 (\partial_\alpha \phi)^2
\end{eqnarray}
and for the tensor part we can have
\begin{equation}
\tilde{P}_{\alpha\beta} =
(\kappa - \epsilon \psi - \beta\psi^2) 
\partial_\alpha \phi \partial_\beta \phi.
\end{equation}
